---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
``` {r global_options, warning=FALSE, message=FALSE}
```

```{r echo=FALSE, results="hide"}
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(scales)
library(lattice)
options(scipen=7)
```
## Loading and preprocessing the data  
(1) Load the data  
``` {r load the data}
df <- read.csv("activity.csv")
```  
(2) Process/transform
``` {r process/transform}
df <- df %>% 
        mutate(date = ymd(date)) %>%
        mutate(interval = str_pad(interval, 4, pad="0")) %>%
        mutate(interval = sub("(^\\d\\d)", "\\1:", interval)) %>%
        mutate(interval = parse_date_time(interval, "%H:%M"))
head(df)
```


## What is mean total number of steps taken per day?

(1) The total number of steps taken per day is as shown in the following chart, for the first five days, at least:  
``` {r steps per day 1}
dfp <- df %>% 
        filter(!is.na(steps)) %>% 
        group_by(date) %>%
        summarize(total_steps = sum(steps))
print(head(dfp))
```  
(2) A histogram showing this data is as follows:  
```{r histogram 1}
qplot(data = dfp, x = total_steps, xlab = "Number of daily steps")
```  

(3) The average number of steps taken per day is `r as.integer(mean(dfp$total_steps))` and the median is `r quantile(dfp$total_steps, probs = 0.5)` as can be seen in the summary table below.  

Mean:  
```{r mean}
as.integer(mean(dfp$total_steps))
```  
Median:  
``` {r median}
quantile(dfp$total_steps, probs = 0.5)
```  
Summary Table:  
``` {r summary table}
summary(dfp)
```

## What is the average daily activity pattern?  
(1) The time series plot of the average number of steps taken by time interval across the number of days is as follows:  
``` {r time series of daily steps}
dfs <- df %>% 
        filter(!is.na(steps)) %>% 
        group_by(interval) %>% 
        summarize(avg_steps=mean(steps))
qplot(data = dfs, x = interval, y = avg_steps, geom = "line", 
      ylab = "Average number of steps per day", 
      xlab = "Time interval") + 
        scale_x_datetime(labels = date_format("%H:%M"))
```  

(2) The time interval containing the maximum number of steps is `r dfs %>% filter(avg_steps==max(avg_steps)) %>% mutate(interval = format(interval, "%H:%M")) %>% select(interval)`.
``` {r interval with max steps}
dfs %>% 
        filter(avg_steps == max(avg_steps)) %>% 
        mutate(interval = format(interval, "%H:%M")) %>% 
        select(interval)
```  

## Imputing missing values  
(1) The number of missing values in the dataset is `r nrow(filter(df, is.na(steps)))`.
``` {r missing values}
nrow(filter(df, is.na(steps)))
```  
(2) To impute the missing values, I will apply the mean for the time of day across all days.  
```{r impute missing values}
dfn <- left_join(df, dfs, by=c("interval"="interval")) %>%
        do(mutate(., steps = ifelse(is.na(steps), avg_steps, steps))) %>%
        select(-avg_steps)
```  
[See here for usage of "do" in mutate](https://github.com/hadley/dplyr/issues/464)  
(3) Create a histogram of the total number of steps taken each day.
``` {r steps by day 2}
dfp2 <- dfn %>% 
        group_by(date) %>%
        summarize(total_steps = sum(steps))
print(head(dfp2))
```

A histogram showing this data is as follows:  
```{r histogram 2}
qplot(data = dfp2, x = total_steps, 
      xlab = "Number of daily steps")
```

## Are there differences in activity patterns between weekdays and weekends?  
(1) Create a variable for weekdays vs. weekends  
```{r weekends vs. weekdays}
dfl <- df %>% 
        filter(!is.na(steps)) %>% 
        mutate(weekday = ifelse(wday(date, label = TRUE) %in% c("Sat", "Sun"), "weekend", "weekday")) %>%
        group_by(interval, weekday) %>%
        summarize(mean_steps = mean(steps))
```  
(2) Panel plot of time series across weekdays or weekends  
``` {r panel plot}
xyplot(mean_steps ~ interval | weekday, 
       data=dfl, layout=c(1,2), type="l")
```
